---
title: "COVID-19 Project"
output: pdf_document
date: "2026-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r visible setup, message=FALSE}
library(tidyverse)
cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")
deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")
```

# Introduction

The five most populous counties in California (Los Angeles, Orange, Riverside, San Bernardino, and San Diego) are all located in Southern California. Despite being geographically contiguous, the counties are dissimilar in a number of ways, from population size to median income and educational attainment (see table below). Given the known effects of certain demographic details like age and income on COVID incidence and mortality, I wanted to investigate whether these counties had notably different outcomes regarding case numbers and mortality rates during the pandemic. 


| County | Population (in millions) | Median Age | Median Household Income | Percentage of Pop. with a Bachelor's Degree or Higher |
|---------------|---------------|---------------|---------------|-----------------|
| **Los Angeles** | 10.0 | 38.9 | \$90,845 | 37.2% |
| **Orange** | 3.2 | 40.2 | \$115,229 | 44.7% |
| **Riverside** | 2.4 | 37.6 | \$93,172 | 26.2% |
| **San Bernardino** | 2.2 | 35.3 | \$88,294 | 24.1% |
| **San Diego** | 3.3 | 37.8 | \$109,132 | 45.2% |

: Data Source: U.S. Census Bureau

## Data

The data used for this project comes from the COVID-19 data repository by the Center for Systems Science and Engineering at Johns Hopkins University. In particular, I used the time series summaries for confirmed cases and deaths in the United States, which includes data from the beginning of the pandemic, in January of 2020, to the point at which Johns Hopkins stopped aggregating data, in March of 2023. All of the case and death totals originally came from each county's department of public health.


# Data Cleaning

There were several columns in the datasets that I did not need in my analysis, such as the Iso codes, latitude, and longitude. As a result, I deleted those. Similarly, I narrowed the data down to the rows that contained information about my five counties of interest. Finally, I pivoted the data in order to have each date represent a row and moved the case and death counts to a column.

After the data was cleaned, I combined the two data sets into one and added columns for deaths per thousand and cases per thousand in order to standardize the case and death counts across the various populations. I also changed the datatype of the values in the date column to dates rather than characters. 

```{r data cleaning}
deaths <- deaths %>%
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Country_Region, -Lat, -Long_, -Combined_Key) %>%
  filter(deaths$Province_State == "California" & (deaths$Admin2 == "San Bernardino" |  
           deaths$Admin2 == "Riverside" | deaths$Admin2 == "San Diego" |  
           deaths$Admin2 == "Los Angeles" | deaths$Admin2 == "Orange")) %>%
  pivot_longer(cols = -c(Admin2, Province_State, Population), 
               names_to = "date", 
               values_to = "deaths")


cases <- cases %>%
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Country_Region, -Lat, -Long_, -Combined_Key) %>%
  filter(cases$Province_State == "California" & (cases$Admin2 == "San Bernardino" |  
           cases$Admin2 == "Riverside" | cases$Admin2 == "San Diego" |  
           cases$Admin2 == "Los Angeles" | cases$Admin2 == "Orange")) %>%
  pivot_longer(cols = -c(Admin2, Province_State), 
               names_to = "date", 
               values_to = "cases")


cases_and_deaths <- full_join(deaths, cases) %>%
    mutate(deaths_per_thousand = deaths/(Population/1000),
           cases_per_thousand = cases/(Population/1000),
           date = mdy(date))
```

# Visualizations

```{r cases graph}
cases_and_deaths %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=cases_per_thousand, color=Admin2)) +
  geom_line() +
  theme_bw() +
  labs(title="Covid Cases in Southern California", 
       x="Date", 
       y="Cumulative Cases (Per Thousand)", color="County") +
  scale_color_discrete(breaks=c("Los Angeles", "San Bernardino", "San Diego", "Riverside", "Orange"))
```

The first thing I wanted to explore with visualizations was the cumulative case counts for the five counties over the time period for which we have data. All of the counties saw spikes in case counts around the beginning of the year in 2021 and 2022. From 2021 to 2022, San Bernardino, Los Angeles, and Riverside counties had greater case counts than did San Diego and Orange counties. By the beginning of 2022, however, Orange county had notably lower cumulative cases than the other four counties did.

```{r deaths graph}
cases_and_deaths %>%
  filter(deaths > 0) %>%
  ggplot(aes(x=date, y=deaths_per_thousand, color=Admin2)) +
  geom_line() +
  theme_bw() +
  labs(title="Covid Deaths in Southern California", 
       x="Date", 
       y="Cumulative Deaths (Per Thousand)", color="County") +
  scale_color_discrete(breaks=c("San Bernardino", "Los Angeles", "Riverside", "Orange", "San Diego"))
```

When it came to COVID deaths, the cumulative totals diverged more starkly In particular, San Diego County, despite having the third highest cumulative case rate, had the lowest number of deaths for almost the entire period for which we have data.  Conversely, Orange County had the fourth highest number of deaths, beginning in 2021, despite having the lowest number of cumulative cases over nearly the same time period.

# Modelling

San Diego and Orange counties are similar in many ways: they have similar median income, educational attainment, and population totals. Thus, I was interested in further exploring the disparity between their pandemic outcomes and seeing if I could see more specifically where their fortunes may have diverged. I used a linear model to predict deaths per thousand, using the cases per thousand information from all of the five counties. Then, I graphed the cases per thousand relative to the deaths per thousand for both counties and added the linear model to the graph for comparison. 

```{r model}
socal_model <- lm(deaths_per_thousand ~ cases_per_thousand, data=cases_and_deaths)

cases_and_deaths <- cases_and_deaths %>%
  mutate(pred = predict(socal_model))

cases_and_deaths %>%
  filter(cases_and_deaths$Admin2 == "Orange" | cases_and_deaths$Admin2 == "San Diego") %>%
  ggplot(aes(x=cases_per_thousand, y=deaths_per_thousand, color=Admin2)) +
  geom_line() +
  geom_line(aes(x=cases_per_thousand, y=pred), color="lightblue3") +
  theme_bw() +
  labs(title="Predicted vs. Actual Covid Deaths",
       x="Cases Per Thousand",
       y="Deaths Per Thousand",
       color="County")

```

# Conclusions

San Diego and Orange counties had somewhat similar mortality rates, relative to their case count, until their respective cumulative cases reached approximately 70 cases per thousand, a benchmark which they both reached in January of 2021. At that point, their trajectories diverged. In the end, Orange County's mortality rate ended up being similar to that predicted by the linear model, while San Diego's remained significantly lower.

There are many potential factors that could have caused this disparity to happen. Orange County's population is notably older than San Diego County's is, with a median age of 40.2 vs. 37.8, and age has been shown to affect COVID mortality rates. 

Additionally, vaccines became widely available in 2021, and vaccine rates may have differed between the two counties. San Diego County is home to multiple military facilities, and members of the military not only qualified for vaccines earlier than the general public did but also were required to get fully vaccinated for COVID (until January of 2023). 

Finally, the political makeup of San Diego and Orange counties differs. According to the California Secretary of State's website, 41.58% and 26.82% of San Diego County residents are registered as Democrats and Republicans, respectively, while those values are 37.61% and 33.07% for Orange County. Since vaccine adoption in the United States has been shown to be correlated with membership in a political party (with Democrats more likely to get vaccinated than Republicans), this disparity in political affiliation could very well have affected vaccine uptake in both counties as well. 

Without taking into account more information about San Diego and Orange counties, it is ultimately impossible to say for sure what led to such disparate outcomes for the two areas. However, all of the factors mentioned above would provide interesting avenues for future inquiry.

# Sources of Bias

First, to acknowledge personal bias: I am a resident of Los Angeles County and resided in that county for the entirety of the time studied here. As such, I saw certain events on the news, like anti-lockdown protests in Huntington Beach (a town in Orange County), which may have affected my views of the five counties studied. However, I made an attempt to be led by the data at all times and to not let my personal biases affect my work.

COVID data was notoriously difficult to source. Many people with COVID either did not get tested for it (early in the pandemic) or used home tests, the results of which they did not report (later in the pandemic). Thus, the case counts likely represent a vast under count of the actual cases of COVID circulating at any time. Additionally, the COVID death metric could be more complicated than it appears. Many people that died with COVID had other comorbidities that contributed to their deaths. It's possible that the death count also is an under count, as COVID might not have been listed as the cause of death on some death certificates, even if it was a factor. 

```{r}
sessionInfo()
```
